---
title: "Comparing Encoder-Based Models for Event Coding^[Prepared for the Advancing the Use of Computational Tools in Political Science Virtual Research Group, held under the auspices of the 2025 American Political Science Virtual Research Meetings. The authors would like to thank Aravind Chinnari for his help with earlier versions of the models presented in this paper. Our code and data can be accessed at https://github.com/eteitelbaum/code-satp.]"
subtitle: "Evidence from the South Asia Terrorism Portal"
author:
  - name: Emmanuel Teitelbaum^[George Washington University, Department of Political Science]
  - name: Mohiddin Bacha Shaik^[George Washington University, Data Science Program]
  - name: Shiva Chakrabarty Sharma^[George Washington University, Trachtenberg School of Public Policy]
date: today
date-format: long
abstract: |
  \vspace{4em}
  \noindent 
  \begin{center} 
  \textit{Early draft. Do not circulate.} \\
  \vspace{4em}
  \textbf{Word count}: 3,710
  \end{center}
  \newpage{}
  \begin{center} 
  \textbf{ABSTRACT}
  \end{center}
  \vspace{1em}
  \noindent \normalsize
  This paper compares the performance of encoder-based transformer models on a series of conflict-related coding tasks. Using hand-coded data from the South Asia Terrorism Portal (SATP) pertaining to the Maoist insurgency in India, we evaluate six models—BERT, RoBERTa, ELECTRA, DistilBERT, XLNet, and the domain-specific ConfliBERT—across multiple dimensions. Events are annotated with overlapping labels capturing forms of violence and state response, allowing us to assess model performance across label frequency, lexical complexity, and training data availability. We also evaluate inference speed to reflect practical constraints in low-resource academic settings. While ConfliBERT performs well overall, particularly on rare and semantically nuanced labels, RoBERTa matches its accuracy on many tasks and runs at comparable speed. All models perform strongly on frequent, lexically distinctive labels but struggle with rare or adjacent categories. These results highlight the promise of general-purpose open-source models for many annotation tasks, the modest gains from domain-specific pretraining, and the persistent challenges of rare label prediction. We conclude with recommendations for improving classification performance on difficult categories and discuss the broader tradeoffs between model complexity, annotation cost, and workflow design in applied conflict research.
  \newpage
#word-count: "8,453"
number-sections: true
format:
  pdf:
    #documentclass: article
    toc: false
    fontsize: 12pt
    mainfont: Palatino
    mathfont: STIX Two Math
    linestretch: 2
    link-citations: true
    colorlinks: true
    linkcolor: Blue4
    geometry: margin=1in
    fig-width: 8
    fig-height: 4
    include-in-header: header.tex
    keep-tex: true
    #block-headings: false
execute: 
  echo: false
  warning: false
  error: false
params: 
  reps: 5000
#reference-section-title: References
bibliography: nlp-coding.bib
#thanks: We are grateful to IPSOS and the GW Politics Poll for fielding the surveys analyzed in this paper.
---

```{r}
#| label: setup
#| include: false

# Load necessary packages
library(ggplot2)
library(dplyr)
library(tidyr)
library(pROC)
library(knitr)
library(viridis)
library(readr)
library(here) 
library(janitor)
library(patchwork)

# Declare location of manuscript
here::i_am("satp-classification.qmd")

# Load custom functions
source("data-viz/model_performance_plots.R")
source("data-viz/heatmap_plots.R")
source("data-viz/support_plots.R")
source("data-viz/scatter_plots.R")
source("data-viz/bootstrap_multiclass_macro_f1_plot.R")
source("data-viz/bootstrap_multiclass_weighted_f1_plot.R")
source("data-viz/bootstrap_multilabel_f1_plot.R")

# Load and mildly wrangle summary data
perpetrator_summary <- read_csv("results/perpetrator_summary.csv") |>
  clean_names()
action_type_summary <- read_csv("results/actiontype_test1_summary.csv") |>
  clean_names() 
target_type_summary <- read_csv("results/targettype_test2_summary.csv") |>
  clean_names()

# Load predictions
perpetrator_pred <- read_csv("results/perpetrator_predictions.csv")
action_type_pred <- read_csv("results/actiontype_test1_predictions.csv") 
target_type_pred <- read_csv("results/targettype_test2_predictions.csv")
```

## Introduction

Large language models (LLMs) are rapidly changing how political scientists and conflict researchers analyze unstructured text. Transformer-based architectures now dominate natural language processing (NLP), offering improved performance over traditional methods such as bag-of-words models, dictionaries, and supervised classifiers. These advances have generated enthusiasm for the possibility that LLMs might dramatically reduce the labor involved in annotating political texts, particularly for tasks like sentiment classification, ideology scaling, and structured event coding.

In political science, early evaluations of general-purpose models such as GPT-3 and GPT-4 demonstrated strong performance on relatively straightforward tasks, especially in English and on short texts. For example, @heseltine2024 show that GPT-4 achieves high accuracy on sentiment and political relevance classification across multiple countries, while @ornstein2023 find that GPT-3 can match or exceed human coding across several political applications. These studies helped establish the feasibility of LLM-assisted annotation, particularly for binary or categorical tasks involving relatively unambiguous inputs.

Building on this foundation, researchers in conflict studies have begun developing domain-specific models trained on political violence and protest corpora. ConfliBERT [@hu2022; @brandt2024] is the most prominent example, a BERT-based model pre-trained on tens of gigabytes of conflict-related text. It consistently outperforms general-purpose models on event classification, actor extraction, and named entity recognition. Similarly, @halterman2023 introduce NGEC, a modular transformer-based pipeline for extracting structured political events, emphasizing flexibility across coding ontologies and contexts.

These models have shifted the conversation from whether LLMs are viable for political science to how best to deploy them. In particular, domain-specific pretraining has been presented as a necessary step for achieving state-of-the-art results in applied research. Yet this assumption is often difficult to evaluate in practice, especially for researchers working with constrained resources or seeking to reuse existing open-access models. Moreover, many of the most time-consuming tasks in political event coding—such as identifying common event types or actor roles—are conceptually routine, even if they appear in bespoke datasets.

This paper revisits the promise and limits of domain-specific LLMs in the context of a series of conflict-related classification tasks. We use a subset of event descriptions from the South Asia Terrorism Portal (SATP), focusing on the Maoist insurgency in India. Events were hand-coded by researchers using a multi-label taxonomy capturing event type, actor behavior, and modalities of violence. The labels vary in frequency and complexity, ranging from common, lexically distinctive labels like *killing* to more semantically ambiguous ones such as *infrastructure attack*. While the dataset shares many structural features with other conflict datasets—such as GTD, ICEWS, or Polecat—it also reflects includes some relatively niche labels that were specific to the interests of the researchers working on this specific project.

Our study makes several contributions. First, we evaluate and compare several transformer-based models commonly used in political text classification: BERT, DistilBERT, ELECTRA, RoBERTa, XLNet, and the domain-specific ConfliBERT. We examine how model performance varies across labels and training data fractions. Second, we incorporate inference speed into our evaluation. This is an important consideration in academic and low-resource environments where compute budgets are constrained. We find that while ConfliBERT performs best overall, RoBERTa matches its performance on many labels and runs at comparable speed—complicating the argument that domain-specific pretraining is always necessary for strong performance.

Third, we assess how models handle lexically complex and low-frequency labels. These are often the very tasks that would be delegated to graduate student coders or domain experts due to their ambiguity or theoretical salience. Our results show that even well-trained models struggle with such categories, and that performance gains are uneven across the label space. While transformer-based models can reliably code frequent and lexically grounded labels with minimal supervision, they remain inconsistent on semantically adjacent or underrepresented categories.

In short, our findings suggest that open-source general-purpose models can be highly effective for a wide range of political coding tasks, including many found in conflict datasets. Domain-specific models like ConfliBERT provide additional accuracy, particularly when label definitions are subtle or data are scarce, but their advantages may be incremental rather than wholly transformative. For many applications—particularly those involving routine coding tasks—open models such as RoBERTa may offer a sufficient balance of speed, accuracy, and accessibility.

The remainder of the paper proceeds as follows. We begin by describing the dataset and coding framework. We then outline the modeling approach, evaluation metrics, and experimental design. Results are presented with attention to model comparisons, error patterns, and performance variation by label type. We conclude with a discussion of implications for future annotation strategies, especially in low-resource research settings, and point to concrete paths for improving model performance on rare and ambiguous labels.

## The SATP Maoist Violence Incidents Dataset

The South Asia Terrorism Portal (SATP) is a foundational resource for scholars of political violence in South Asia. Its curated event summaries offer one of the most comprehensive, longitudinal collections of open-source conflict data on the region, with particular relevance for research on insurgency, terrorism, and counterinsurgency in India. Recent work in political event classification using large language models (LLMs) has employed SATP as a source of annotated conflict data for supervised learning and model evaluation. @hu2022 used the SATP dataset to test ConfliBERT, a BERT-based model trained on a large corpus of political violence texts, evaluating it on binary classification tasks like identifying whether a report described a violent event and multi-label classification like assigning event types such as armed assault or bombing. @brandt2024 extended this work by comparing ConfliBERT to more recent decoder-based LLMs, using SATP again for illustrative examples while shifting the primary evaluation focus to larger-scale datasets such as the Global Terrorism Database (GTD).

In our study, we draw on the same underlying SATP news reports to construct a distinct dataset focused exclusively on violence associated with the Maoist insurgency. Our dataset spans the years 2005 to 2016, a period that overlaps with but also precedes the SATP-based annotations used in the ConfliBERT studies. This timeframe was selected to align with the peak intensity of Maoist activity in India. Like Hu et al., we work at the level of incident summaries reported in SATP, but our coding is tailored to a single conflict and employs a more detailed and conflict-specific annotation framework. Our dataset includes approximately 10,000 events and is designed to test whether transformer-based models can effectively classify complex, context-dependent event types grounded in a specific political and geographic setting.

The dataset was constructed through a multi-year coding project involving undergraduate coders trained in political science and South Asian studies, working under the supervision of graduate researchers with domain expertise. Coders relied on a detailed codebook, developed and refined over time, and entered structured data for each incident using a standardized digital form. Each report was carefully parsed to determine whether it contained multiple events or described a single compound incident. Every incident received a unique identifier encoding the state, date, and event sequence for that day.

The coding schema includes a range of variables not commonly found in automated datasets including the perpetrator (or action taker) and a broad range of target and action types. Where information allowed, coders also captured the organizational role of perpetrators (e.g., commander, cadre, or sympathizer), and made distinctions between government and private infrastructure when coding targets. Action types were disaggregated to distinguish, for example, between armed assaults and seizures, or between facility attacks and bombings. These distinctions reflect coding priorities informed by field knowledge and scholarly interest in the heterogeneity of violence within a single conflict.

The coding process was characterized by an emphasis on quality control. To ensure consistency and reliability, coders underwent structured onboarding, participated in supervised trial phases, and engaged in regular review and adjudication sessions to resolve edge cases. Overlapping assignments were used to assess and calibrate inter-coder reliability. A significant share of the dataset was double-coded, with discrepancies reviewed and resolved by senior coders. In cases where coding standards evolved over time, previously coded events were revisited during the quality assurance phase to harmonize labels and ensure consistency across the full time series.

While ConfliBERT and related BERT-based models have shown promising results on classification tasks involving relatively broad event categories, our dataset presents a more varied and nuanced test of LLM capabilities. By coding more finely grained distinctions within a single conflict, we aim to assess how well pretrained models can handle the complexity of political violence data when applied to bespoke annotation schemes. This challenge reflects the reality that researchers working with real-world conflict data often require models that can disaggregate events in ways that reflect local variation, political context, and shifting repertoires of violence. In the following section, we outline the specific tasks we set out to evaluate and compare how our selected models performed on them. 

## Methods

To evaluate how effectively pretrained language models can recover structured political violence data from unstructured text, we implemented a series of classification tasks using our hand-coded SATP Maoist Violence Incidents Dataset. The tasks reflect key dimensions of political violence—actor type, action type, and target type—each operationalized as either multi-class or multi-label classification problems.

We evaluate model performance on three supervised learning tasks:

- A multi-class classification task to identify the *perpetrator* (or actor) involved in each incident (Maoist, Security, or Other).
- A multi-label classification task to identify the *actions* taken, with possible labels including Armed Assault, Bombing, Abduction, Seizure, Arrest, Surrender, and Facility/Infrastructure Attack.
- A second multi-label classification task to identify *targets* of an attack or other action, with categories including Security, Civilians, Officials, Government Infrastructure, Private Property, Mining Companies, NGOs, Other Groups, and Maoists.

These tasks are extractive in nature and require the model to recover structured labels from unstructured incident summaries. For this reason, we restrict our evaluation to encoder-only transformer models, which are better suited to classification and token-level prediction than decoder-only models like GPT-style architectures. Encoder models generate bidirectional contextual embeddings that allow them to attend to all parts of the input simultaneously, a key advantage in tasks that require understanding and resolving multiple co-occurring elements.

We selected six models for evaluation, all freely available via the Hugging Face Model Hub:

- *BERT-base-cased* [@devlin2019]: A 12-layer, 110M parameter model pretrained on BooksCorpus and Wikipedia. It serves as the baseline encoder model for classification tasks.
- *DistilBERT-base-cased* [@sanh2020]: A 6-layer, 66M parameter distilled version of BERT that is faster and lighter, designed to reduce inference cost with minimal performance loss.
- *RoBERTa-base* [@liu2019]: A 12-layer, 125M parameter BERT variant trained with more data and no next sentence prediction, generally outperforming BERT on downstream tasks.
- *ELECTRA-base-discriminator* [@clark2020]: A 12-layer, 110M parameter model that uses a more sample-efficient pretraining objective (replaced token detection) rather than masked language modeling.
- *XLNet-base-cased* [@yang2020]: A 12-layer, 110M parameter autoregressive model trained with permutation-based language modeling, capable of modeling bidirectional context while retaining autoregressive benefits.
- *ConfliBERT* [@hu2022]: A BERT-base model pretrained on a corpus of political violence and conflict-related text. It is domain-specific and designed to improve performance on tasks involving political conflict data.

All six models fall into the "base" category (between 66–125M parameters), making them feasible to run on commodity hardware, including laptops and free or low-tier Google Colab environments. This makes them particularly well suited for use in low-resource settings—both in terms of computational infrastructure and access to large labeled datasets. Their relative efficiency also allows for more flexible experimentation, such as evaluating performance across multiple training data fractions.

Each model was fine-tuned on a fixed training–validation–test split drawn from the full dataset. Approximately 2,000 incident descriptions were randomly selected and held out at the start: 1,000 for validation and 1,000 for final testing. The remaining ~8,000 incidents were used in the training loop, where we iteratively varied the size of the training set across six fractions: $1/32$, $1/16$, $1/8$, $1/4$, $1/2$, and the full training set. Validation and test sets remained fixed throughout, allowing consistent comparison of performance across data fractions and model architectures. All models were fine-tuned using their cased versions, based on the assumption that capitalization conveys important semantic and contextual information in this domain—particularly for identifying proper nouns, such as location names, political figures, or militant organizations, which are common in South Asia–specific reporting.

To evaluate model performance, we report weighted macro F1 scores for the multi-class perpetrator classification task and micro F1 scores for the two multi-label classification tasks (action type and target type). These metrics were selected to account for class imbalance while preserving interpretability and comparability across settings. Additional performance metrics, including precision, recall, and per-label F1 scores, will be provided in an online supplement.

## Results

### Perpetrator Task

The perpetrator task is a multi-class classification problem in which models are trained to predict the actor type responsible for a given incident, with classes including Maoist, Security, and Other. Figure 1a displays the weighted macro F1 scores achieved by each model across different training data fractions. Figure 1b focuses on a single fraction ($1/8$ of the training set), presenting 95% confidence intervals obtained via bootstrap resampling (5,000 iterations) to assess statistical uncertainty around model estimates.

```{r}
#| label: perpetrator
#| fig-cap: "Model Performance for 'Perpetrator' Task"
#| fit-align: center

line_chart <- model_performance_plot(
  data = perpetrator_summary,
  xvar = fraction_raw,
  yvar = eval_f1_weighted,
  color = model_label,
  title = "a. F1 vs. Training Data Fraction",
  ytitle = "Weighted F1 Score",
  palette = "turbo",
  end = .8
)

f1_ci_plot <- bootstrap_multiclass_weighted_f1_plot(perpetrator_pred,
                             fraction_value = "12%",
                             title = "b. CIs at 12% Training",
                             color = "#8337A0",
                             n_boot = params$reps)

combined_plot <- line_chart + f1_ci_plot + 
  plot_layout(widths = c(2, 1)) + 
  plot_annotation(
    #title = "Model Performance for 'Perpetrator' Task",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
    caption = "95% confidence intervals for macro F1 scores calculated via bootstrap resampling."
  )

combined_plot
```
Consistent with findings reported in the original ConfliBERT paper [@hu2022], we observe that ConfliBERT performs particularly well at lower data fractions, where its domain-specific pretraining appears to confer an advantage. However, the gap narrows quickly. At the 1/8 training level, RoBERTa and XLNet both achieve F1 scores that are not statistically distinguishable from ConfliBERT’s, perhaps complicating a straightforward narrative about the superiority of domain-specific model superiority [@brandt2024; @osorio2024].

When trained on the full dataset, all models converge to very high weighted macro F1 scores, ranging from 0.957 (ConfliBERT) to 0.968 (RoBERTa). This tight clustering in performance likely reflects the relative simplicity of the classification task (predicting whether an actor is Maoist, Security, or Other) and the benefit of high-quality, consistent human annotations. While ConfliBERT retains a slight advantage in low-resource settings, the results at full training suggest that general-purpose encoder models are capable of matching or exceeding the performance of domain-specific alternatives on this task. In a later section, we return to inference speed and other performance trade-offs, which further contextualize these results for applications in low-resource environments.

### Action Type Task

The second task is a multi-label classification problem in which models predict one or more action types associated with a given incident. These include Armed Assault, Bombing, Abduction, Seizure, Arrest, Surrender, and Facility/Infrastructure Attack. Performance is evaluated using micro F1 scores, which are better suited to imbalanced, multi-label settings where many samples carry multiple labels or none at all.

Figure 2a presents model performance across increasing training data fractions. As in the perpetrator task, we observe notable variation in learning rates. RoBERTa and XLNet clearly lead at the 25% training level (Figure 2b), with micro F1 scores of 0.882 and 0.879, respectively. ConfliBERT trails modestly behind at 0.857, while BERT and DistilBERT cluster around 0.800 and 0.785. ELECTRA performs substantially worse, achieving a score of only 0.636 at this level.

```{r}
#| label: action_type_performance
#| fig-cap: "Model Performance for 'Action Type' Task"
#| fig-align: left

line_chart <- model_performance_plot(
  data = action_type_summary,
  xvar = fraction_raw,
  yvar = eval_micro_avg_f1_score,
  color = model_label,
  title = "a. F1 vs. Training Data Fraction",
  ytitle = "Micro F1 Score",
  palette = "turbo",
  end = .8
)

f1_ci_plot <- bootstrap_multilabel_f1_plot(action_type_pred,
                             fraction_value = "25.0%",
                             title = "b. CIs at 25% Training",
                             ytitle = NULL,
                             color = "#8337A0",
                             n_boot = params$reps)

combined_plot <- line_chart + f1_ci_plot + 
  plot_layout(widths = c(2, 1)) +
  plot_annotation(
    #title = "Model Performance for 'Action Type' Task",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
    caption = "95% confidence intervals for micro F1 scores calculated via bootstrap resampling."
  )

combined_plot
```
When trained on the full dataset, all models perform well, converging to micro F1 scores in a narrow band between 0.915 (ELECTRA) and 0.939 (XLNet). The top four models—XLNet, RoBERTa, BERT, and ConfliBERT—are separated by only a few hundredths of a point, underscoring a high level of overall convergence. Once again, ConfliBERT performs strongly in low-data regimes but is closely matched or outperformed by general-purpose encoder models when more training data are available.


```{r}
#| label: action_type_labels
#| fig-cap: "Model Performance and Test Set Support by 'Action Type' Label"
#| fig-align: center

library(patchwork)

# Generate the heatmap and store the object
heatmap <- per_label_f1_heatmap(action_type_summary)

# Extract the label order used in the heatmap
label_order <- levels(heatmap$data$label)

# Generate the support bar plot using the same label order
barplot <- per_label_support_barplot(action_type_summary, label_levels = label_order) + 
  theme(axis.text.x = element_text(size = 8))

# Combine with patchwork
combined_plot <- barplot + heatmap +
  plot_layout(widths = c(1, 3)) +
  plot_annotation(
    #title = "Model Performance and Test Set Support by 'Action Type' Label",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
    #caption = "Barplot shows the number of samples per label in the test set. Heatmap shows the F1 score for each model and label."
  )

combined_plot
```
These headline results, however, mask important variation across action types. As shown in Figure 3, some labels (e.g., Surrender) are predicted with very high accuracy despite low support in the training set, while others (e.g., Facility/Infrastructure Attack) perform less well despite being relatively more common. This suggests that label frequency alone does not fully explain model performance. Certain labels may benefit from linguistic distinctiveness—for example, the word “surrender” appears with few synonyms or paraphrases in the source texts—while others, like infrastructure-related actions, may involve greater lexical ambiguity or more diverse expressions, making them harder to classify.

### Target Type Task

The final task involves predicting one or more target types associated with each incident. As in the action classification task, this is a multi-label setting, and performance is again evaluated using micro F1 scores. Target categories include Security, Civilians, Officials, Government Infrastructure, Private Property, Mining Companies, NGOs, Other Groups, and Maoists.

Figure 4a presents model performance across training fractions, with Figure 4b zooming into the 50% training level, where we again observe a clear performance cluster among XLNet, ConfliBERT, and RoBERTa. At this level, XLNet leads with a micro F1 score of 0.839, followed by ConfliBERT (0.819) and RoBERTa (0.816). DistilBERT and BERT score considerably lower (0.770 and 0.710, respectively), and ELECTRA again falls well behind the others at 0.614.

```{r}
#| label: target_type_performance
#| fig-cap: "Model Performance for 'Target Type' Task"
#| fig-align: center

line_chart <- model_performance_plot(
  data = target_type_summary,
  xvar = fraction_raw,
  yvar = eval_micro_avg_f1_score,
  color = model_label,
  title = "a. F1 vs. Training Data Fraction",
  ytitle = "Micro F1 Score",
  palette = "turbo",
  end = .8
)

f1_ci_plot <- bootstrap_multilabel_f1_plot(target_type_pred,
                             fraction_value = "50.0%",
                             title = "b. CIs at 50% Training",
                             color = "#8337A0",
                             n_boot = params$reps)

combined_plot <- line_chart + f1_ci_plot + 
  plot_layout(widths = c(2, 1)) +
  plot_annotation(
    #title = "Model Performance for 'Target Type' Task",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
    caption = "95% confidence intervals for micro F1 scores calculated via bootstrap resampling."
  )

combined_plot
```
At full training (Figure 4a, rightmost points), model scores converge somewhat, but still lag those observed in the action type task. ConfliBERT reaches 0.878, with RoBERTa (0.870) and XLNet (0.869) close behind. However, performance tails off for BERT (0.847), DistilBERT (0.826), and ELECTRA (0.787). These figures indicate that target prediction is a more difficult task overall, likely due to higher class imbalance and greater semantic ambiguity in the expression of target entities.

```{r}
#| label: target_type_labels
#| fig-cap: "Model Performance and Test Set Support by 'Target Type' Label"
#| fig-align: center

library(patchwork)

# Create shorter axis labels
short_labels = c("Maoist", "Security", "Civilians", "Infrastructure", "Property", "Officials", "Mining Co.", "NGOs", "Other Group")

# Generate the heatmap and store the object
heatmap <- per_label_f1_heatmap(target_type_summary) + scale_x_discrete(labels = short_labels)

# Extract the label order used in the heatmap
label_order <- levels(heatmap$data$label) 

# Generate the support bar plot using the same label order
barplot <- per_label_support_barplot(target_type_summary, label_levels = label_order) +
  scale_x_discrete(labels = short_labels) +
  theme(axis.text.x = element_text(size = 8))

# Combine with patchwork
combined_plot <- barplot + heatmap +
  plot_layout(widths = c(1, 3)) +
  plot_annotation(
    #title = "Model Performance and Test Set Support by 'Action Type' Label",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
    #caption = "Barplot shows the number of samples per label in the test set. Heatmap shows the F1 score for each model and label."
  )

combined_plot
```

Figure 5 further illustrates this point by breaking down F1 scores across individual labels. Some target types—particularly those with limited support in the training set—show markedly lower performance. However, the decline is not solely attributable to frequency. For example, Surrender in the action task achieved strong scores despite low support, likely due to its lexical distinctiveness. By contrast, Infrastructure/Facility targets underperform even when relatively frequent, suggesting that greater lexical variability and ambiguity in how these targets are described contributes to model error.

### Performance-Speed Tradeoffs

Figure 6 presents a performance–speed comparison for the action and target type classification tasks, plotting micro F1 scores against evaluation speed (measured in samples per second). The figure highlights the trade-offs between model accuracy and runtime efficiency, which are especially relevant for researchers working in low-resource or latency-sensitive environments.

```{r}
#| label: performance_vs_speed
#| fig-cap: "Model Performance vs. Speed for Multiclass Classification Tasks"
#| fig-align: center

library(grid)

action_scatter <- scatter_plot_speed_vs_accuracy(
  data = action_type_summary,
  xcol = eval_samples_per_second,
  ycol = eval_micro_avg_f1_score,
  color = model_label,
  size = fraction_raw,
  title = "Action Type"
) + 
  xlab(NULL) + 
  theme(
    legend.position = "none",
    panel.border = element_rect(color = "gray80", fill = NA, linewidth = 0.5),
    panel.background = element_blank()
  ) +
  scale_size_continuous(
    range = c(2, 7),
    breaks = c(0.03125, 0.0625, 0.25, 0.50, 0.75, 1.00),
    name = "Fraction Raw"
)

target_scatter <- scatter_plot_speed_vs_accuracy(
  data = target_type_summary,
  xcol = eval_samples_per_second,
  ycol = eval_micro_avg_f1_score,
  color = model_label,
  size = fraction_raw,
  title = "Target Type"
) + 
  labs(x = NULL, y = NULL) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.border = element_rect(color = "gray80", fill = NA, linewidth = 0.5),
    panel.background = element_blank()
  ) +
  scale_size_continuous(
    range = c(2, 7),
    breaks = c(0.03125, 0.0625, 0.25, 0.50, 0.75, 1.00),
    name = "Fraction Raw"
)

# Combine
combined_speed_plot <- (action_scatter + target_scatter) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "right",
    legend.box.margin = margin(0, 0, 0, 0),
    legend.key.size = unit(0.4, "cm"),
    legend.text = element_text(size = 8),
    plot.margin = margin(5, 5, 5, 5)
  )

# Add title/caption (but not x-axis label)
combined_speed_plot <- combined_speed_plot +
  plot_annotation(
    #title = "Model Performance vs. Speed for Multiclass Classification Tasks",
    #caption = "The size of the points indicates the fraction of data used for training.",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),  # affects only the annotation title
      plot.caption = element_text(margin = margin(t = 10))
    )
  )

# Draw plot with shared x-axis label using grid
#grid.newpage()
grid.draw(combined_speed_plot)
grid.text("Evaluation Samples Per Second", y = unit(0.02, "npc"), gp = gpar(fontsize = 12))
```
Among the top-performing models, RoBERTa and ConfliBERT stand out for their efficiency. At full training, both achieve micro F1 scores near the top of the range—0.870 and 0.878, respectively (on the target task)—while maintaining high evaluation speeds of 147 and 145 samples per second. BERT, while slightly less performant (0.847), operates at a comparable speed (144 samples/sec), making it a viable option where pretraining data diversity is sufficient.

In contrast, XLNet, though highly competitive in terms of accuracy (0.869), is much slower, evaluating at only 65 samples per second—less than half the speed of RoBERTa and ConfliBERT. This substantial performance-cost trade-off limits XLNet’s practical appeal in time-constrained settings.

DistilBERT emerges as the fastest model, running at 267 samples per second, but as seen earlier, it tends to underperform at lower training fractions, making it less suitable for small-data scenarios. ELECTRA performs poorly on both axes: it is the least accurate model (0.787) and also slower than most of its peers (124 samples/sec), suggesting limited utility for either high-accuracy or speed-critical tasks.

Though Figure 6 focuses on the action and target type tasks, patterns are similar for the perpetrator classification task (not shown). In all cases, the results suggest that RoBERTa and ConfliBERT offer the best balance between accuracy and efficiency, particularly when high throughput is required alongside strong classification performance.

## Conclusion

This study evaluated the performance of large language models (LLMs) on a multi-label classification task using a subset of hand-coded events from the South Asia Terrorism Portal (SATP) covering the Maoist insurgency in India. The dataset reflects the type of bespoke coding often found in social science research, where events are labeled across multiple dimensions with a custom taxonomy developed for a specific analytical purpose.

Our analysis showed that LLMs performed reasonably well on high-frequency labels, particularly when those labels were lexically distinctive or closely tied to surface-level textual features. For instance, labels like killing and arrest were predicted with relatively high accuracy even with limited training data. As the volume of training data increased, model performance improved overall, but the gains were not evenly distributed across labels.

Performance on low-frequency labels remained limited, with models often failing to predict these categories reliably. Some labels, such as infrastructure attacks, saw improvements with more training data, but still lagged behind more common categories. We also observed that multi-label performance varied by model and training fraction, suggesting that LLMs are sensitive to both model capacity and data availability in these settings.

These findings indicate that LLMs can serve as effective tools for assisting with annotation on certain types of labels—especially those that are frequent and lexically grounded—but remain less reliable for rare or semantically diffuse labels. In practice, this means that LLMs could potentially reduce the human labor required for annotating well-represented event types, while human oversight is still needed for more challenging cases.

Future work should explore strategies for improving performance on the tail of the label distribution. One possibility is oversampling rare labels during training to mitigate class imbalance. Another is data augmentation, including paraphrasing or targeted generation of examples for underrepresented categories. More structured approaches, such as organizing labels hierarchically or embedding them in a continuous space, may also help models generalize across related labels.

In addition, further analysis of model errors could inform refinements to both the coding scheme and the annotation process. For example, identifying which labels are frequently confused may help clarify category definitions or guide the construction of auxiliary tasks to sharpen model distinctions.

Finally, while this study focused on a single dataset, the challenges it raises are broadly relevant to social scientists considering the use of LLMs for classification tasks. Bespoke annotation schemes and imbalanced label distributions are common features of applied research, and this work contributes initial guidance on where LLMs are likely to be useful and where further development is needed. The SATP dataset provides a valuable testbed for continued experimentation with these methods.

## References 